The Hyperedge Event Model

### 2018-05-24 paper/ba-HEM/ba-sample.tex
revisions until page 17 of BA draft

In recent decades, real-time digitized textual communication has developed into a ubiquitous form of social and professional interaction \citep{kanungo2008modeling, szostek2011dealing, burgess2004email, pew2016}. From the perspective of the computational social scientist, this has led to a growing need for methods of modeling interactions that manifest as edges exchanged in continuous time. A number of models that build upon topic modeling through Latent Dirichlet Allocation \citep{Blei2003} to incorporate link data as well as textual content have been developed recently \citep{mccallum2005author,lim2013twitter,Krafft2012}. These models are innovative in their extensions that incorporate network information. However, none of the models that are currently available in the literature integrate the rich random-graph structure offered by state of the art models for network structure---such as the exponential random graph model (ERGM) \citep{robins2007introduction,chatterjee2013estimating,hunter2008ergm}. The ERGM is the canonical model for modeling the structure of a static network. It is flexible enough to specify a generative model that accounts for nearly any pattern of edge formation such as reciprocity, clustering, popularity effects \citep{desmarais2017statistical}.

Several network models have been developed that handle time-stamped events in which edge formation is governed by structural dynamics similar to those used in the ERGM \citep{Butts2008,Vu2011,snijders1996stochastic}. Those models are very useful to understand which traits and behaviours are predictive of interactions, however, none of the models explicitly allows the hyperedge---a connection between two or more vertices generated simultaneously\footnote{A hyperedge connecting just two nodes is simply a usual dyadic edge.}---which is a common property of digitalized textual interactions such as emails and online messages. For instance, \cite{PerryWolfe2012} treat multicast interactions---one type of directed hyperedge which involve one sender and multiple receivers---via duplication (i.e., obtain pairwise interactions from the original multicast), but their model parameter estimation relies on likelihood approximation which leads to bias. Similarly, \cite{fan2009learning} treats multicast emails as multiple single-receiver edges and randomly jitter the sent times, to avoid the violation of continous-time model assumption. On the other hand, there have been numerous studies for hyperedges under the hypergraph theory \citep{karypis1999multilevel} in phyics literature, including a random hypergraph model \citep{ghoshal2009random}, but most of them are designed specifically for engineering applications---e.g., multiple tagging networks \citep{zlatic2009hypergraph,zhang2010hypergraph}, cellular networks \citep{klamt2009hypergraphs}, and personalized recommendations \citep{zhang2010personalized,blattner2009b}. To our knowledge, there is a lack of extensive efforts to incorportate hyperedge modeling into general statistical framework.

Concentrating on developing a statistical network model with better handling of hyperedges, we introduce the hyperedge event model (HEM) which simultaneously models the two components that govern time-stamped event formation: 1) the receiver selection processs that allows multiple senders or receivers, and 2) the time-to-next interactions with flexible distributional choices. In what follows, we introduce the HEM by describing how we assume the generatve process of a time-stamped event data (Section \ref{sec:generative process}), and deriving the sampling equations for Bayesian inference (Section \ref{sec:inference}). Then, we apply the model to the Montgomery county government email data and perform two model validation tasks (Section \ref{subsec:Emails}). Finally, the paper finishes in Section 5 with conclusion and discussion.
\section{The Hyperedge Event Model}\label{sec:generative process}

Data generated under the model consists of $D$ unique edges. A single edge, indexed by $d \in [D]$, is represented by the three components: the sender $a_d \in [A]$, an indicator vector of receivers $\boldsymbol{r}_d = \{u_{dr} \}_{r=1}^{A}$, and the timestamp $t_d \in (0, \infty)$. For simplicity, we assume that edges are ordered by time such that $t_d \leq t_{d+1}$. While the model can be applied for two type of hyperedges---edges with (1) one sender and one or more receivers, and (2) one receiver and mone or more receivers---here we only present the generative process for those involving one sender and one or more receivers (i.e., multicast). For the latter case of hyperedges, we treat $a_d$ to be an indicator vector of senders $\boldsymbol{a}_d = \{u_{dr} \}_{a=1}^{A}$ and $r_d$ to be the single receiver. Detailed generative process for directed edges with one receiver and one or more senders are provided in Appendix A.

\subsection{Hypothetical Edges}\label{subsec: Tie}
For every possible sender--receiver pair $(a,r)_{a \neq r}$, we define the ``receiver intensity"---an approximate inverse logit of the probability that edge $d$ is being sent from the sender $a$ to receiver $r$---as a linear combination of network-based statistics:
\begin{equation}
\lambda_{adr} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{adr},
\end{equation}
where $\boldsymbol{b}$ is a $P$--dimensional vector of coefficients and $\boldsymbol{x}_{adr}$ is a set of receiver selection features which vary depending on the hypotheses regarding canonical processes relevant to network theory such as popularity, reciprocity, and transitivity. In addition, we include intercept term to account for the average (or baseline) number of receivers. We place a Normal prior $\boldsymbol{b} \sim N(\boldsymbol{\mu}_b, \Sigma_b)$.

Next, we hypothesize ``If $a$ were the sender of edge $d$, who would be the receivers?" and derive a receiver-selection distribution. For an edge $d$, we first define an $A\times A$ matrix $\boldsymbol{u}_d$ where the $a^{th}$ row denotes sender $a$'s receiver vector of indicators---i.e., if node $r$ is the hypothetical receiver of sender $a$, then $u_{adr}=1$; otherwise $u_{adr}=0$. We then assume that each receiver vector $\boldsymbol{u}_{ad}$ comes from the multivariate Bernoulli (MB) distribution \citep{dai2013multivariate}---a model to estimate the structure of graphs with binary nodes---with logit probability of $\boldsymbol{\lambda}_{ad}$. In order to avoid the model degeneracy from having an empty receiver set, we define a probability measure ``MB$_{G}$" motivated by the non-empty Gibbs measure \citep{fellows2017removing} which excludes the all-zero vector from the support of the multivariate Bernoulli distirbution. As a result, this measure helps us to 1) allow multiple sender--receiver pairs to co-exist, 2) force the sender to select at least one receiver, and 3) ensure tractable normalizing constant. To be specific, we draw a binary vector $\boldsymbol{u}_{ad}= (u_{ad1},
\ldots, u_{adA})$ 
\begin{equation} \boldsymbol{u}_{ad}  \sim
\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad}),
\end{equation}
where $\boldsymbol{\lambda}_{ad}= \{\lambda_{adr}\}_{r=1}^A$. In particular, we define $\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad})$ as
\begin{equation}
\begin{aligned}
&\Pr(\boldsymbol{u}_{ad}|\boldsymbol{b}, \boldsymbol{x}_{ad}) = \frac{1}{Z(\boldsymbol{\lambda}_{ad})}\exp\Big(\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ad}\rVert_1 > 0 )\big) + \sum_{r\neq a} \lambda_{adr}u_{adr}\Big) ,
\end{aligned}
\label{eqn:Gibbs}
\end{equation}
where $Z(\boldsymbol{\lambda}_{ad})= \prod_{r \neq a} \big(\mbox{exp}(\lambda_{adr}) + 1\big)-1$ is the normalizing constant and $\lVert \cdot \rVert_1$ is the $l_1$--norm. Again, this is equivalent to assuming independent Bernoulli trial on each $u_{adr}$ with probability of 1 being logit($\lambda_{adr}$), excluding the case when $u_{adr}=0$ for all $r \in [A]$. We provide the derivation of the normalizing constant as a tractable form in Appendix B.

\subsection{Hypothetical Timestamps}\label{subsec:Time}
Similarly, we consider ``If $a$ were the sender of edge $d$, when would it be sent?" and define the ``timing rate" for the sender $a$
\begin{equation}
\mu_{ad} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ad}),
\end{equation}
where $\boldsymbol{\eta}$ is a $Q$--dimensional vector of coefficients with a Normal prior $\boldsymbol{\eta} \sim N(\boldsymbol{\mu}_\eta,\Sigma_\eta)$, $\boldsymbol{y}_{ad}$ is a set of event timing features, e.g., any covariates that could affect timestamps of the edge, and $g(\cdot)$ is the appropriate link function such as identity, log, or inverse.

In modeling ``when," we do not directly model the timestamp $t_d$. Instead, we assume that each sender's ``time increment"---i.e., waiting time to next interaction---is drawn from a specific distribution in the exponential family. We define the time increment from edge $d-1$ to edge $d$ as $\tau_{d}$ (i.e., $\tau_{d}= t_d-t_{d-1}$) and specify the distribution of hypothetical timestamps with sender-specfic mean $\mu_{ad}$. Following the generalized linear model (GLM) framework \citep{nelder1972generalized}, we assume the mean and variance of the $\tau_{ad}$ satistify
\begin{equation}
\begin{aligned}
E(\tau_{ad}) &= \mu_{ad},\\
V(\tau_{ad}) &= V(\mu_{ad}),
\end{aligned}
\end{equation}
where $\tau_{ad}$ here is a positive real number. Possible choices of distribution include exponential, Weibull, gamma, and log-normal\footnote{The log-normal distribution is not exponential family but can be used via modeling of $\log(\tau_d)$.} distributions, which are commonly used in time-to-event modeling \citep{rao2000applied,rizopoulos2012joint}. Based on the choice of distribution, we may introduce any additional latent variable (e.g., variance parameter $\sigma_\tau^2$ for log-normal and shape parameter $\theta$ for gamma) to account for the variance in time increments. We use $f_\tau(\cdot; \mu)$ and $F_\tau(\cdot; \mu)$ to denote the probability density function (p.d.f) and cumulative density function (c.d.f), respectively, with mean $\mu$.
\subsection{Senders, Receivers, and Timestamps}\label{subsec:Observed}
Finally, we choose the actual sender, receivers, and timestamp by selecting the sender--receiver-set pair with the smallest time increment \citep{snijders1996stochastic}:
\begin{equation}
\begin{aligned}
a_d &= \mbox{argmin}_{a}(\tau_{ad}),\\
\boldsymbol{r}_d &= \boldsymbol{u}_{a_d d},\\
t_d &=t_{d-1} + \tau_{a_d d}.
\end{aligned}
\end{equation}
Therefore, it is a sender-driven process in that the receivers and timestamp of an edge is jointly determined by the sender's urgency to send the edge to chosen receivers. Note that our generative process accounts for tied events such that in case of tied events (i.e., multiple senders generated exactly same time increments), we observe all of the tied events without assigning the orders of tied events. Algorithm \ref{alg:generative} summarizes the entire generative process for directed edges with one sender and one or more receivers, outlined in this section.
	\begin{algorithm}[!t]
		\spacingset{1}
			\SetAlgoLined
		\caption{Generative Process: one sender and one or more receivers}
		\begin{algorithmic}
				\STATE \textbf{Input}: number of edges and nodes $(D, A)$, covariates $(\boldsymbol{x}, \boldsymbol{y})$, and the coefficients $(\boldsymbol{b}, \boldsymbol{\eta})$
				\vskip 0.1in
			\FOR{d=1 to D}
			\FOR{a=1 to $A$}
			\FOR{r=1 to $A$ (r $\neq$ a)}
			\STATE	set $\lambda_{adr} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{adr}$
			\ENDFOR
			\STATE	draw $\boldsymbol{u}_{ad}  \sim
			\mbox{MB}_G(\boldsymbol{\lambda}_{ad})$
			\STATE		set $\mu_{ad} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ad})$
			\STATE		draw $\tau_{ad} \sim f_\tau(\mu_{ad}, V(\mu_{ad}))$
			\ENDFOR

@@

Unlike hypothetical edges above, new values for $\boldsymbol{b}$ cannot be sampled directly from its conditional posterior, but may instead be obtained using the Metropolis--Hastings algorithm. Assuming an uninformative prior (i.e., $N({0},\infty)$), the conditional posterior over $\boldsymbol{b}$ is

@@

Likewise, we use Metropolis-Hastings algorithm to update the latent variable $\boldsymbol{\eta}$. Assuming an uninformative prior $\boldsymbol{\eta}$ (i.e., $N({0},\infty)$), the conditional posterior for no-tied event case is

@@

where $t_1^*,\ldots,t_M^*$ are the unique timepoints across $D$ events ($M \leq D$). If $M=D$ (i.e., no tied events), equation (\ref{eqn:latenttime2}) reduces to equation (\ref{eqn:latenttime}). Note that when we have the latent variable to quantify the variance in time increments $V(\cdot)$ (based on the choice of timestamp distribution in Section \ref{subsec:Time}), we also use equation (\ref{eqn:latenttime}) (or equation (\ref{eqn:latenttime2}) in case there exist tied events) for the additional Metropolis-Hastings update---e.g., $\Pr(\sigma^2_\tau| \boldsymbol{\eta},\boldsymbol{u}, \boldsymbol{y},\boldsymbol{a}, \boldsymbol{r},\boldsymbol{t})$ for log-normal distribution and $\Pr(\theta| \boldsymbol{\eta},\boldsymbol{u}, \boldsymbol{y},\boldsymbol{a}, \boldsymbol{r},\boldsymbol{t})$ for gamma distribution.

@@

					\STATE update $u_{adr}$ using Gibbs update --- equation (\ref{eqn:latentreceiver})

@@

			\STATE update $\boldsymbol{b}$ using Metropolis-Hastings --- equation (\ref{eqn:latentedge})

@@

			\STATE update $\boldsymbol{\eta}$ using Metropolis-Hastings --- equation (\ref{eqn:latenttime}) or equation (\ref{eqn:latenttime2}) 
			\STATE (if needed) update the latent variable for $V(\cdot)$ using Metropolis-Hastings --- equation (\ref{eqn:latenttime}) or equation (\ref{eqn:latenttime2}) 

@@

	\ENDFOR
	\STATE	summarize the results with the last chain of $\boldsymbol{b}$ and $\boldsymbol{\eta}$
\end{algorithmic}
\label{alg:MCMC}
\end{algorithm}
\subsection{Getting It Right (GiR) Test} \label{subec:GiR}
  Software development is integral to the objective of applying the HEM to real world data. Code review is a valuable process in any research computing context, and the prevalence of software bugs in statistical software is well documented \citep[e.g., ][]{altman2004numerical,mccullough2009accuracy}.  With highly complex models such as the HEM, there are many ways in which software bugs can be introduced and go unnoticed. As such, we present a joint analysis of the integrity of our generative model, sampling equations, and software implementation.

@@

   In the forward samples, both observable and unobservable variables are generated using Algorithm \ref{alg:generative}. In the backward samples, unobservable variables are generated using the sampling equations for inference, which we derived in Section \ref{subsec:conditionaldist}. For each forward and backward sample that consists of $D$ number of edges, we save these statistics:
   \begin{itemize}
   	\item[1.] Mean of observed receiver sizes $ \lVert \boldsymbol{r}_{d} \rVert_1 $ across $d=1,\ldots,D$,
   	\item[2.] Mean of time increments $\tau_d$ across $d=1,...,D$,
   	\item[3.] $b_p$ value used to generate the samples $p = 1,...,P$,
   	\item[4.] $\eta_q$ value used to generate the samples $q = 1,...,Q$,
   	\item[5.] $\sigma^2_\tau$ value used to generate the samples in log-normal distribution
   \end{itemize}

@@

 	   we formulate the edge covariates $\boldsymbol{x}$ and timestamp covariates $\boldsymbol{y}$ and ground them in illustrative examples. We then report a suite of experiments that test the HEM's ability to form the posterior distribution over latent variables. 

@@

 	   	\includegraphics[width=0.65\textwidth]{img/triad-1.png}	

@@

	The right plot of Figure \ref{figure:PPEresults} presents boxplots for the time prediction errors. where plot the estimates in a log-scale. Surprisingly, we have a huge benefit in the performance of timestamp prediction when we assume log-normal distribution for time-increments compared to exponential distribution. This difference can be explained by overdispersion in exponential distribution, because there exists greater variability in the time increments of emails than would be expected under exponential distribution. As illustrated above, we can use this out-of-sample prediction task for two usage---1) provide an effective answer to the question ``how are we filling in the missing components of time-stamped network data?" and 2) offer one standard way to determine the distribution of time increments in Section \ref{subsec:Time}. 

@@

	\subsection{Posterior Predictive Checks}\label{subsec:PPC_email} 	   
In this section, we perform posterior predictive checks (PPC) \citep{rubin1984bayesianly} to evaluate the appropriateness of our model specification for Montgomery County email data. We formally generated entirely new data, by simulating $N=500$ synthetic email data $\{(a_{d}, \boldsymbol{r}_{d}, t_{d})\}_{d=1}^D$ from the genenerative process in Section \ref{sec:generative process}, conditional upon a set of inferred latent variables from the inference in Section \ref{subsec:Result_email}, where the pseudocode is outlined in Algorithm \ref{alg:PPC}. For the test of goodness-of-fit in terms of network dynamics, we use multiple statistics that summarize meaningful aspects of the data: outdegree distribution, indegree distribution, receiver size distribution, and probability-probability (PP) plot for time increments. 
		\caption{Generate new data for PPC}
			\STATE \textbf{Input}: number of data to generate $R$, covariates $(\boldsymbol{x},\boldsymbol{y})$, and estimated values of $(\boldsymbol{u}, \boldsymbol{b}, \boldsymbol{\eta})$\\

			\FOR{$r=1$ {\bfseries to}  $R$}
			\FOR{$d = 1$ {\bfseries to}  $D$} 
			\STATE	Draw ($a_{d}$, $\boldsymbol{r}_{d}$, $t_{d}$) following Algorithm \ref{alg:generative}\\
			\ENDFOR
			\STATE Store every $r^{th}$ new data $\{(a_{d}, \boldsymbol{r}_{d}, t_{d})\}_{d=1}^D$ 
			\ENDFOR
		\end{algorithmic}
		\label{alg:PPC}
	\end{algorithm}

>>>>

In recent decades, real-time digitized textual communication has developed into a ubiquitous form of social and professional interaction \citep{kanungo2008modeling, szostek2011dealing, burgess2004email, pew2016}. From the perspective of the computational social scientist, this has led to a growing need for methods of modeling interactions that manifest as edges exchanged in continuous time. In recent decades, several dynamic network models which could handle time-stamped events with temporal granularity have been developed, where their edge formations are governed by structural dynamics similar to those used in the exponential random graph model (ERGM)---the canonical model for modeling the structure of a static network. For example, stochastic actor-oriented models (SAOMs) \citep{snijders1996stochastic,snijders2007modeling} study the network evolutions as a consequence of \textit{micro-steps}---i.e., actors making choices to form a new tie or withdraw an existing tie. On the other hand, event-based network models \citep{Butts2008,Vu2011,hunter2011dynamic} provide a general regression framework for longitudinal networks by modeling an edge as a realization of continuous time survival process. Those models are flexible enough to specify a generative model that accounts for nearly any pattern of edge formation such as reciprocity, clustering, popularity effects \citep{desmarais2017statistical}, thus very useful to understand which traits and behaviours are predictive of interactions.

However, most studies above are limited to modeling edges with one sender and one receivers. To our knowledge, none of the existing continuous-time network models in statistical literature explicitly allow hyperedges---a connection between two or more vertices generated simultaneously\footnote{A hyperedge connecting just two nodes is simply a usual dyadic edge.}---which is a common property of digitalized textual interactions such as emails and online messages. For instance, \cite{PerryWolfe2012} treat multicast interactions---one type of directed hyperedge which involve one sender and one or more receivers---via duplication (i.e., obtain pairwise interactions from the original multicast), to construct approximate likelihood function in their inferential framework for model parameters. Similarly, \cite{fan2009learning} duplicates emails sent from one sender to one or more receivers and randomly jitter the sent times, in order to avoid the violation of continous time model assumption. There have been numerous studies for hyperedges under the hypergraph theory \citep{karypis1999multilevel} in phyics literature, including a random hypergraph model \citep{ghoshal2009random}, but most of them are designed specifically for engineering applications---e.g., multiple tagging networks \citep{zlatic2009hypergraph,zhang2010hypergraph}, cellular networks \citep{klamt2009hypergraphs}, and personalized recommendations \citep{zhang2010personalized,blattner2009b}---with lack of extensive efforts to incorportate hyperedge modeling into general statistical framework.

Concentrating on developing a statistical network model with better handling of hyperedges, we introduce the hyperedge event model (HEM) which jointly models the two components that govern time-stamped event formation: 1) the receiver selection processs that allows directed edges with one sender and one or more receivers or one receiver and one or more senders, and 2) the timing equation that models time increments (i.e., time to next event) with flexible distributional choices. In what follows, we introduce the HEM by describing how we assume the generatve process of a time-stamped event data (Section \ref{sec:generative process}), and deriving the conditional posteriors for Bayesian inference along with code review tests (Section \ref{sec:inference}). Then, we demonstrate the model's applicability by analyzing one of the county government email network---the Montgomery county email data---via model selection, posterior predictive checks, and exploratory analysis (Section \ref{subsec:Emails}). Finally, the paper finishes in Section 5 with conclusion and discussion.
\section{The Hyperedge Event Model}\label{sec:generative process}

Data generated under the model consists of $D$ unique edges. A single edge, indexed by $d \in [D]$, is represented by the three components: a sender $a_d \in [A]$, an indicator vector of receivers $\boldsymbol{r}_d = \{u_{dr} \}_{r=1}^{A}$, and a timestamp $t_d \in (0, \infty)$. For simplicity, we assume that edges are ordered by time such that $t_d \leq t_{d+1}$. While the model can be applied for two type of hyperedges---edges with (1) one sender and one or more receivers, and (2) one receiver and mone or more receivers---here we only present the generative process for those involving one sender and one or more receivers (i.e., multicast). For the latter case of hyperedges, we treat $a_d$ to be an indicator vector of senders $\boldsymbol{a}_d = \{u_{dr} \}_{a=1}^{A}$ and $r_d$ to be the single receiver. Detailed generative process for directed edges with one receiver and one or more senders are provided in Appendix A.

\subsection{Hypothetical Edges}\label{subsec: Tie}
For every possible sender--receiver pair $(a,r)_{a \neq r}$, we define the ``receiver intensity"---an approximate inverse logit of the probability that edge $d$ is being sent from sender $a$ to receiver $r$---as a linear combination of statistics relevant to the receiver selection process:
\begin{equation}
\lambda_{adr} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{adr},
\end{equation}
where $\boldsymbol{b}$ is a $P$--dimensional vector of coefficients and $\boldsymbol{x}_{adr}$ is a set of receiver selection features which vary depending on the hypotheses regarding canonical processes relevant to network theory such as popularity, reciprocity, and transitivity. In addition, we include intercept term to account for the average (or baseline) number of receivers. We place a Normal prior $\boldsymbol{b} \sim N(\boldsymbol{\mu}_b, \Sigma_b)$.

Next, we hypothesize ``If $a$ were the sender of edge $d$, who would be the receivers?" and derive a receiver-selection distribution. For an edge $d$, we first define an $A\times A$ matrix $\boldsymbol{u}_d$ where the $a^{th}$ row denotes sender $a$'s receiver vector of indicators---i.e., if node $r$ is the hypothetical receiver of sender $a$, then $u_{adr}=1$; otherwise $u_{adr}=0$. We then assume that each receiver vector $\boldsymbol{u}_{ad}$ comes from the multivariate Bernoulli (MB) distribution \citep{dai2013multivariate}---a model to estimate the structure of graphs with binary nodes---with logit probability of $\boldsymbol{\lambda}_{ad}$. In order to avoid the model degeneracy from having an empty receiver set, we define a probability measure ``MB$_{G}$" motivated by the non-empty Gibbs measure \citep{fellows2017removing} which excludes the all-zero vector from the support of the multivariate Bernoulli distirbution. As a result, this measure helps us to 1) allow multiple sender--receiver pairs to co-exist, 2) force the sender to select at least one receiver, and 3) ensure tractable normalizing constant. To be specific, we draw a binary vector $\boldsymbol{u}_{ad}= (u_{ad1},
\ldots, u_{adA})$ 
\begin{equation} \boldsymbol{u}_{ad}  \sim
\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad}),
\end{equation}
where $\boldsymbol{\lambda}_{ad}= \{\lambda_{adr}\}_{r=1}^A$. In particular, we define $\mbox{MB}_{G}(\boldsymbol{\lambda}_{ad})$ as
\begin{equation}
\begin{aligned}
&\Pr(\boldsymbol{u}_{ad}|\boldsymbol{b}, \boldsymbol{x}_{ad}) = \frac{1}{Z(\boldsymbol{\lambda}_{ad})}\exp\Big(\mbox{log}\big(\text{I}( \lVert \boldsymbol{u}_{ad}\rVert_1 > 0 )\big) + \sum_{r\neq a} \lambda_{adr}u_{adr}\Big) ,
\end{aligned}
\label{eqn:Gibbs}
\end{equation}
where $Z(\boldsymbol{\lambda}_{ad})= \prod_{r \neq a} \big(\mbox{exp}(\lambda_{adr}) + 1\big)-1$ is the normalizing constant and $\lVert \cdot \rVert_1$ is the $l_1$--norm. Again, this is equivalent to assuming independent Bernoulli trial on each $u_{adr}$ with probability of 1 being logit($\lambda_{adr}$), excluding the case when $u_{adr}=0$ for all $r \in [A]$. We provide detailed derivation steps for the normalizing constant in Appendix B.

\subsection{Hypothetical Timestamps}\label{subsec:Time}
Similarly, we consider ``If $a$ were the sender of edge $d$, when would it be sent?" and define the ``timing rate" for sender $a$
\begin{equation}
\mu_{ad} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ad}),
\end{equation}
where $\boldsymbol{\eta}$ is a $Q$--dimensional vector of coefficients with a Normal prior $\boldsymbol{\eta} \sim N(\boldsymbol{\mu}_\eta,\Sigma_\eta)$, $\boldsymbol{y}_{ad}$ is a set of event timing features, e.g., any covariates that could affect timestamps of the edge, and $g(\cdot)$ is the appropriate link function such as identity, log, or inverse.

In modeling ``when," we do not directly model the timestamp $t_d$. Instead, we assume that each sender's ``time increment"---i.e., waiting time to next interaction---is drawn from a specific distribution in the exponential family. We define the time increment from edge $d-1$ to edge $d$ as $\tau_{d}$ (i.e., $\tau_{d}= t_d-t_{d-1}$) and specify the distribution of hypothetical timestamps with sender-specfic mean $\mu_{ad}$. Following the generalized linear model (GLM) framework \citep{nelder1972generalized}, we assume the mean and variance of the $\tau_{ad}$ satistify
\begin{equation}
\begin{aligned}
E(\tau_{ad}) &= \mu_{ad},\\
V(\tau_{ad}) &= V(\mu_{ad}),
\end{aligned}
\end{equation}
where $\tau_{ad}$ here is a positive real number. Possible choices of distribution include exponential, Weibull, gamma, and log-normal\footnote{The log-normal distribution is not exponential family but can be used via modeling of $\log(\tau_d)$.} distributions, which are commonly used in time-to-event modeling \citep{rao2000applied,rizopoulos2012joint}. Based on the choice of distribution, we may introduce any additional latent variable (e.g., shape parameter $k$ for Weibull, shape parameter $\theta$ for gamma, and variance parameter $\sigma_\tau^2$ for log-normal) to account for the variance in time increments. We use $f_\tau(\cdot; \mu, V(\mu))$ and $F_\tau(\cdot; \mu, V(\mu))$ to denote the probability density function (p.d.f) and cumulative density function (c.d.f), respectively, with mean $\mu$ and variance $V(\mu)$.
\subsection{Senders, Receivers, and Timestamps}\label{subsec:Observed}
Finally, we choose the actual sender, receivers, and timestamp by selecting the sender--receiver-set pair with the smallest time increment \citep{snijders1996stochastic}:
\begin{equation}
\begin{aligned}
a_d &= \mbox{argmin}_{a}(\tau_{ad}),\\
\boldsymbol{r}_d &= \boldsymbol{u}_{a_d d},\\
t_d &=t_{d-1} + \tau_{a_d d}.
\end{aligned}
\end{equation}
Therefore, it is a sender-driven process in that the receivers and timestamp of an edge is jointly determined by the sender's urgency to send the edge to chosen receivers. Note that our generative process accounts for tied events such that in case of tied events (i.e., multiple senders generated exactly same time increments), we observe all of the tied events without assigning the orders of tied events. Algorithm \ref{alg:generative} summarizes the entire generative process for directed edges with one sender and one or more receivers, outlined in this section.
	\begin{algorithm}[!t]
		\spacingset{1}
			\SetAlgoLined
		\caption{Generative Process: one sender and one or more receivers}
		\begin{algorithmic}
				\STATE \textbf{Input}: number of edges and nodes $(D, A)$, covariates $(\boldsymbol{x}, \boldsymbol{y})$, and the coefficients $(\boldsymbol{b}, \boldsymbol{\eta})$
				\vskip 0.1in
			\FOR{d=1 to D}
			\FOR{a=1 to $A$}
			\FOR{r=1 to $A$ (r $\neq$ a)}
			\STATE	set $\lambda_{adr} = {\boldsymbol{b}}^{\top}\boldsymbol{x}_{adr}$
			\ENDFOR
			\STATE	draw $\boldsymbol{u}_{ad}  \sim
			\mbox{MB}_G(\boldsymbol{\lambda}_{ad})$
			\STATE		set $\mu_{ad} = g^{-1}(\boldsymbol{\eta}^\top \boldsymbol{y}_{ad})$
			\STATE		draw $\tau_{ad} \sim f_\tau(\mu_{ad}, V(\mu_{ad}))$
			\ENDFOR

@@

Unlike hypothetical edges above, new values for $\boldsymbol{b}$ cannot be sampled directly from its conditional posterior, but may instead be obtained using the Metropolis--Hastings (M-H) algorithm. Assuming an uninformative prior (i.e., $N({0},\infty)$), the conditional posterior over $\boldsymbol{b}$ is

@@

Likewise, we use the M-H algorithm to update the latent variable $\boldsymbol{\eta}$. Assuming an uninformative prior $\boldsymbol{\eta}$ (i.e., $N({0},\infty)$), the conditional posterior for no-tied event case is

@@

where $t_1^*,\ldots,t_M^*$ are the unique timepoints across $D$ events ($M \leq D$). If $M=D$ (i.e., no tied events), equation (\ref{eqn:latenttime2}) reduces to equation (\ref{eqn:latenttime}). Note that when we have the latent variable to quantify the variance in time increments $V(\mu)$ (based on the choice of timestamp distribution in Section \ref{subsec:Time}), we also use equation (\ref{eqn:latenttime}) (or equation (\ref{eqn:latenttime2}) in case there exist tied events) for the additional M-H update---e.g., $\Pr(k| \boldsymbol{\eta},\boldsymbol{u}, \boldsymbol{y},\boldsymbol{a}, \boldsymbol{r},\boldsymbol{t})$ for Weibull, $\Pr(\theta| \boldsymbol{\eta},\boldsymbol{u}, \boldsymbol{y},\boldsymbol{a}, \boldsymbol{r},\boldsymbol{t})$ for gamma, and  $\Pr(\sigma^2_\tau| \boldsymbol{\eta},\boldsymbol{u}, \boldsymbol{y},\boldsymbol{a}, \boldsymbol{r},\boldsymbol{t})$ for log-normal distribution.

@@

					\STATE update $u_{adr}$ using Gibbs update ---equation (\ref{eqn:latentreceiver})

@@

			\STATE update $\boldsymbol{b}$ using M-H algorithm---equation (\ref{eqn:latentedge})

@@

			\STATE update $\boldsymbol{\eta}$ using M-H algorithm---equation (\ref{eqn:latenttime}) or (\ref{eqn:latenttime2}) 

@@

			\IF {extra parameter for $V(\mu)$} 
			\STATE update the variance parameter using M-H algorithm---equation (\ref{eqn:latenttime}) or (\ref{eqn:latenttime2}) 
				\ENDIF
	\STATE	summarize the results with the last chain of $\boldsymbol{b}$ and $\boldsymbol{\eta}$
\end{algorithmic}
\label{alg:MCMC}
\end{algorithm}
\subsection{Getting It Right (GiR) Test} \label{subec:GiR}
  Software development is integral to the objective of applying the HEM to real world data. Code review is a valuable process in any research computing context, and the prevalence of software bugs in statistical software is well documented \citep[e.g., ][]{altman2004numerical,mccullough2009accuracy}.  With highly complex models such as the HEM, there are many ways in which software bugs can be introduced and go unnoticed. As such, we present a joint analysis of the integrity of our generative model, sampling equations, and software implementation.

@@

   In the forward samples, both observable and unobservable variables are generated using Algorithm \ref{alg:generative}. In the backward samples, unobservable variables are generated using the sampling equations for inference, which we derived in Section \ref{subsec:conditionaldist}. For each forward and backward sample that consists of $D$ number of edges, we save these statistics:
   \begin{itemize}
   	\item[1.] Mean of observed receiver sizes $ \lVert \boldsymbol{r}_{d} \rVert_1 $ across $d=1,\ldots,D$,
   	   	\item[2.] Variance of observed receiver sizes $ \lVert \boldsymbol{r}_{d} \rVert_1 $ across $d=1,\ldots,D$,
   	\item[3.] Mean of time increments $\tau_d$ across $d=1,...,D$,
   	   	\item[4.] Variance of time increments $\tau_d$ across $d=1,...,D$,
   	\item[5.] $b_p$ value used to generate the samples $p = 1,...,P$,
   	\item[6.] $\eta_q$ value used to generate the samples $q = 1,...,Q$,
   	\item[7.] $\sigma^2_\tau$ value used to generate the samples in log-normal distribution
   \end{itemize}

@@

 	   we formulate the edge covariates $\boldsymbol{x}$ and timestamp covariates $\boldsymbol{y}$ and ground them in illustrative examples. We then report a suite of experiments that test the HEM's ability to form the posterior distribution over its latent variables. 

@@

 	   	\includegraphics[width=0.6\textwidth]{img/triad-1.png}	

@@

@@

	The right plot of Figure \ref{figure:PPEresults} presents boxplots for the median absolute relative errors. where plot the estimates in a log-scale. Surprisingly, we have a huge benefit in the performance of timestamp prediction when we assume log-normal distribution for time-increments compared to exponential distribution. This difference can be explained by overdispersion in exponential distribution, because there exists greater variability in the time increments of emails than would be expected under exponential distribution. As illustrated above, we can use this out-of-sample prediction task for two usage---1) provide an effective answer to the question ``how are we filling in the missing components of time-stamped network data?" and 2) offer one standard way to determine the distribution of time increments in Section \ref{subsec:Time}.

	\subsection{Posterior Predictive Checks}\label{subsec:PPC_email} 	   
In this section, we perform posterior predictive checks (PPC) \citep{rubin1984bayesianly} to evaluate the appropriateness of our model specification for Montgomery County email data. We formally generated entirely new data, by simulating $N=500$ synthetic email data $\{(a_{d}, \boldsymbol{r}_{d}, t_{d})\}_{d=1}^D$ from the genenerative process in Section \ref{sec:generative process}, conditional upon a set of inferred latent variables from the inference in Section \ref{subsec:Result_email}\iffalse, where the pseudocode is outlined in Algorithm \ref{alg:PPC}\fi. For the test of goodness-of-fit in terms of network dynamics, we use multiple statistics that summarize meaningful aspects of the data: outdegree distribution, indegree distribution, receiver size distribution, and probability-probability (PP) plot for time increments. 
\iffalse
	\begin{algorithm}[!t]
		\spacingset{1}
		\caption{Generate new data for PPC}
		\begin{algorithmic}
			\STATE \textbf{Input}: number of data to generate $R$, covariates $(\boldsymbol{x},\boldsymbol{y})$, and estimated values of $(\boldsymbol{u}, \boldsymbol{b}, \boldsymbol{\eta})$\\
			\vskip 0.1in
